---
title: "TChroNetR Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{TChroNetR_Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)


```{r setup}
library(TChroNetR)
```
# Introduction

This vignette demonstrates how to use TChroNetR to build threshold-based chromatin networks, detect communities, select optimal parameters, and perform downstream analyses such as genomic annotation and target gene identification.

# 1. Building a TChroNet Series Object
A TChroNet Series object contains network metrics computed across multiple thresholds. By setting `run_cd = TRUE`, community detection (Leiden algorithm) is performed automatically for each resolution specified in `resolutions_list`.

```{r}
# List all graph edge files (thresholded networks)
edge_files <- list.files("~/T-ChroNet/toy_data/results/th/", full.names = TRUE)

# Path to the normalized count matrix
matrix_path <- "~/T-ChroNet/toy_data/data/thp1_test.tsv"

# Build the TChroNet series object
obj_series <- build_TChroNetSeries_object(
  edge_files,
  matrix_path,
  resolutions_list = seq(0.5, 1.5, 0.1),
  run_cd = TRUE,
  seed = 1234,
  transitivity = FALSE
)
```
# 2. Selecting the Best Threshold

Determine the optimal threshold using the `find_best_th` function. By default, selection is based on the relative largest component. If transitivity was computed, it can also be used as a criterion (note: may slow down series building).
```{r}
# Automatically determine the best threshold parameter
obj_series <- find_best_th(obj_series)
```
# 3. Visualizing Community Detection Results

Explore community structures across thresholds and resolutions:
## Random index map among neighbors
```{r}
plot_randindex_map(obj_series)
```
## Sankey plot at a fixed threshold
```{r}
plot_community_sankey(obj_series, threshold = 0.8)
```
## Sankey plot at a fixed resolution
```{r}
plot_sankey_fixed_resolution(obj_series, resolution = 0.8)
```
# 4. Building a TChroNet Object

Convert the series object into a TChroNet Object for downstream analysis:
```{r}
# Build the final TChroNet object
obj <- build_TChroNetNetwork_from_series(obj_series, threshold = 0.8)
```
Or, if the desired threshold is already known, build directly:
```{r}
# Directly build TChroNet object without scanning thresholds
obj <- make_TChroNetNetwork_obj(
  edge_files,
  matrix_path,
  threshold = 0.8
  )
```
# 5. Community Detection Across Resolutions

If the TChroNet object lacks community detection:
```{r}
# Run community detection at multiple resolutions
obj <- compute_membership_nodes(obj, resolutions = seq(0.5, 1.5, 0.1))
```
Select the best resolution:
## Modularity plot
```{r}
plot_modularity(obj)
```
## Sankey plot to track community stability
```{r}
plot_community_sankey(obj)
```
## Automatically find the best resolution
```{r}
obj <- find_best_resolution(obj)
```
# 6. Visualizing Community Trends

Plot trends of communities:
```{r}
# Plot trends as lines 
plot_line_trends(obj)

# Plot trends as violin
plot_trends_zscore(obj)
```

# 7. Genomic Annotation

Annotate genomic regions. If network coordinates are in hg19, use liftOver to convert to hg38:
```{r}
# Lift hg19 to hg38 if required
obj <- lift_network_coordinates(obj, chain_path = "~/T-ChroNet/toy_data/data/hg19ToHg38.over.chain")

# Annotate with SCREEN cCRE database
obj <- annotate_regions_from_bed(obj, bed_path =  "~/T-ChroNet/toy_data/data/GRCh38-cCREs.bed")
```
# Visualize genomic annotations
```{r}
genomic_position_stackbar(obj)
```

# 8. Target Gene Identification

Identify potential target genes for each community using rGREAT:
```{r}
# Run GREAT functional analysis on the communities
obj <- run_GREAT_analysis(obj, genome = 'hg19')
```


# 9. Transcription Factors Identification